XRONOS: An Open Data Infrastructure for Archaeological Chronology

### 2025-04-04 paper.tex
Rerender as preprint
Incorporate @nevrome's suggestions
Did what I could with fig-c14-datasets-map... the problem is that adding
extra elements (i.e. scales) causes patchwork to squeeze the plot area
even more.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  number,
  doubleblind]{elsarticle}

@@

\affiliation[1]{organization={University of Bern, Institute of
Archaeological Sciences},,postcodesep={}}
\affiliation[2]{organization={Max Planck Institute for
Geoanthropology},,postcodesep={}}

@@

In the remainder of this paper, we describe the concept and
implementation of XRONOS in relation to the state of the art in open
chronometric data in archaeology, and evaluate our progress in achieving
these goals as of writing. Since we envisage both XRONOS as a dataset
and XRONOS as software to be continually developing resources, the
description here should be read as a `snapshot' of the project as of
writing rather than its final state.

@@

The number of available compilations has increased exponentially since
around 1995 (Figure~\ref{fig-c14-datasets-time}). The first generation
came around the turn of the century and consists mostly of online
databases with a web frontend. These include some databases operated by
radiocarbon labs, for example the Oxford Radiocarbon Lab (ORAU) and the
Belgian Royal Institute for Cultural Heritage (KIK-IRPA), and
essentially represent a continuation of their date lists in a digital
format. The majority, however, were compiled from the literature by
individual researchers interested in a particular region and/or period.
Notable early examples include ANDES 14C in 1994 (Central Andes,
Michczyński et al., 1995), CARD (Canada, Gajewski et al., 2011) and
RADON (Europe, Raetzel-Fabian, 1999) in 1999, and CANEW in 2001 (Near
East, Reingruber and Thissen, 2005). From 2010, coinciding with broader
shifts in scientific publishing (Tenopir et al., 2011), it became more
common to publish standalone `open data' products in the form of journal
supplements, archives in repositories and/or data papers; the
\emph{\href{https://openarchaeologydata.metajnl.com}{Journal of Open
Archaeology Data}}, launched in 2012, has been a prominent venue for
this latter category. Most recently there has been a trend towards
providing version-controlled plain text data via platforms such as
\href{https://github.com}{GitHub}, reflecting the broader adoption of
these tools amongst computational archaeologists over the last decade
(Batist and Roe, 2024). The shift from online databases towards more
static but more preservable open data products is welcome, given how
many databases from the first generation have subsequently ceased to be
accessible. Version-controlled repositories are particular well-suited
to data compilation projects because they allow for continued updates
whilst still providing snapshot `releases' that are citeable and can be
archived in long-term repositories.

@@

\caption{\label{fig-c14-datasets-map}Geographic coverage of published
regional radiocarbon compilations according to our survey (see
Supplementary Material).}

@@

\caption{\label{fig-c14-global}Geographic and temporal (sum calibration)
of georeferenced dates in XRONOS and other global radiocarbon
compilations}

@@

Radiocarbon has been by far the most active area of open data
compilation, but archaeological chronology incorporates a much more
diverse range of sources of information (Harding, 1999). In periods
beyond the practical limit of radiocarbon dating (c.~55,000 BP), other
types of radiometric (K--Ar, U--Pb, etc.), chemical or luminescence
dating offer an alternative (Aitken, 1999). Conversely, in historic
periods, radiocarbon is often relatively underused compared to
conventional typological dating (based on artefact characteristics),
which in these periods can offer comparable or better temporal
resolution, or direct dating based on epigraphy (Heřmánková et al.,
2021), numismatics (Kemmers and Myrberg, 2011) or historical sources. In
places where it is widely available, dendrochronology (Baillie, 2014)
also produces significantly better resolved chronologies and therefore
tends to be the main source of chronometric data. Other more
application-specific chronological methods include shoreline dating
(Brøgger, 1905; Roalkvam, 2023), lichenometry (Benedict, 2009) and rock
weathering dating (Bednarik, 2020; Whitley, 2012).

@@

DendroDB, hosted by the Swiss Federal Institute for Forest, Snow and
Landscape Research (WSL), emphasizes ecological and climate studies over
archaeological wood material. While it claims a broad scope, the
database remains non-functional, rendering it ineffective for research
needs. The CFS-TRenD database, managed by the Canadian Forest Service,
compiles over 4,600 datasets from Canadian forests, primarily focusing
on boreal ecosystems. Despite its extensive coverage for North America,
its geographical specificity and lack of open access restrict its
utility for European prehistoric contexts. Similarly, the QUB
Dendrochronology Database, managed by Queen's University Belfast, offers
valuable datasets for Ireland and the UK but lacks significant
representation of prehistoric material, limiting its application in
broader archaeological investigations. The Building Archaeology Research
Database (BARD) contains over 24,000 records, including
dendrochronological data from more than 2,700 buildings. However, its
focus on medieval and post-medieval timber-frame construction further
narrows its utility for studies involving prehistoric wood samples.

@@

Typological dates---i.e.~relative, expertise- or seriation-based dating
based on artefact characteristics---are ubiquitous in archaeological
studies but rarely treated as a form of chronometric data in their own
right. For example, the majority of the radiocarbon datasets we reviewed
(Section~\ref{sec-c14-compilation}) included some form of typological
chronological information in the form of a `period' or `culture' column.
This is also typically present in many other forms of systematic
compilation work in archaeology, for example site gazetteers. Aggregated
typological information from such sources are often used in aoristic
analysis and related methods (Crema, 2024; Mischka, 2004). What is
lacking in this presentation of typological dating is metadata on how
the determination was made and how exactly it is to be understood. Like
any archaeological date, a typological date is derived from a physical
sample -- the object or set of object from which a chronological
estimate was derived. Typological dates on one class of object may well
clash with other classes of object, or for that matter with scientific
dates -- does one trust the date on pottery, the date on architecture,
or the radiocarbon date? Without additional metadata on e.g.~who made
the typological determination or what the radiocarbon date was obtained
on, such inconsistencies are difficult to resolve Similarly the absolute
date range corresponding to a typological determination (e.g.~``Late
Neolithic'') can be interpreted in multiple ways depending on the region
and intentions of the expert making the determination. PeriodO
(Rabinowitz et al., 2016) is a linked open data infrastructure that
includes a shared vocabulary of typological periods and corresponding
calendar age estimates, and an important step towards addressing the
latter problem. However, it remains to be systematically linked to
compilations of typological dates (though there are some efforts in this
direction e.g. Hannah et al., 2022).

@@

XRONOS is our answer to Kintigh's call (Kintigh, 2006) for digital
infrastructures that don't just provide access to chronological data but
enables researchers to ``archive, access, integrate, and mine disparate
data sets''. It complements several similar open data infrastructures
within and outwith archaeology, such as the Global Biodiverisity
Information Facility (GBIF, Canhos et al., 2004), the Strategic
Environmental Archaeology Database (SEAD, Buckland, 2014), IMPACT for
mummified human remains (Nelson and Wade, 2015), Neotoma for
palaeoecological data (Williams et al., 2018), IsoArcH for stable
isotope data (Plomp et al., 2022), the International Soil Radiocarbon
Database (ISRaD, Lawrence et al., 2020), and the `Big Interdisciplinary
Archaeological Database' (BIAD), an ambitious new initiative to combine
many of these individual domains, including chronology (Reiter et al.,
2024). To improve upon existing global syntheses of radiocarbon dates
(see Section~\ref{sec-global-compilations}), we aimed to develop a
living infrastructure that both continually collected data from diverse
sources and presented a seamless single database to the user.

@@

Chrons are conceptually useful because they emphasise that different
types of archaeological `dates', drawn from different sources, have
essentially the same information content: the location of an event in
space and time. We thereby avoid privileging certain sources of
chronological data over (as might be the case if, for example, we
treated `period' as a fixed attribute of a site) and can accommodate
contradictory (e.g.~differences of opinion on typological
classification). This is important given that XRONOS aspires to be an
authoratative `backbone' with a global scope, so we cannot realistically
impose a single chronological scheme or resolve conflicting information
provided by specialists. They are useful practically because they expose
a common interface for attributes that all types of chronological
information share, such as a \emph{terminus post quem} (TPQ),
\emph{terminus ante quem} (TAQ), and midpoint estimate. This allows
applications that use XRONOS' data model (including XRONOS itself) to
collate chronological data from multiple sources, without necessarily
having to be aware of the pecularities of each type of dating.

@@

The series of relations
\texttt{{[}chron{]}\ \textgreater{}\ sample\ \textgreater{}\ context\ \textgreater{}\ site}
links the chronological and contextual sides of the XRONOS data model.
Each step is a many-to-one association, meaning for example that it is
possible to attach multiple chrons to the same sample (e.g.~replicated
radiocarbon dates on the same material), multiple \emph{types} of chrons
to the same sample (e.g.~radiocarbon dates on tree-rings for
wiggle-matching). Since this kind of information is rarely
systematically recorded in our source databases, there are currently few
actual records that make use of this feature of the data model. However,
we hope it will provide a foundation for more nuanced chronological
modelling in the future.

@@

The XRONOS web application exposes two distinct user interfaces: a
graphical user interface accessed through a web browser; and an
application programming interface (API). Both interfaces follow a REST
(Representational State Transfer) pattern (Verborgh et al., 2015), where
each resource (e.g.~a single radiocarbon date, a single bibliographic
reference, or a single user) is statelessly mapped to a single address.
Users can then interact with resources at these addresses using a
preditable and uniform interface based on HTTP verbs. For example, the
radiocarbon date RTD-8904 is represented by the address
\url{https://xronos.ch/c14s/156205}. Users can view information on this
resource by sending a GET request to that address, regardless of which
interface they are using, and authorised users can modify it using POST,
PATCH, or DELETE requests. The bibliographic reference associated with
this date (Richter et al., 2017) is similarly represented at the address
\url{https://xronos.ch/references/17778}, and can be accessed at that
address using the same interface as the radiocarbon date. These uniform
REST interfaces are another example of a boring architectural choice
that make it easier for us to enrich the scientific contents of XRONOS,
by adding new types of modular resource that represent new scientific
entities.

This basic REST pattern is augmented by seven `actions' (following the
standard pattern in Rails application) that express different ways of
interacting with a resource: index, show, destroy, new, create, edit,
and update. The `show' action represents interaction with a single
resource, as described above. The `index' action, which lists resources
of a given type (e.g.~\url{https://xronos.ch/c14s} for radiocarbon
dates), is worth special mention because it is through this that the
filtering logic at the core of XRONOS' two interfaces is implemented. By
passing a query as HTTP GET parameters to the index action of a
resource, the list returned the user is modified to only include records
that match that query. For example,
\url{https://xronos.ch/sites?site\%5Bcountry_code\%5D=CH} (the part of
the URL after the \texttt{?} character encodes the SQL WHERE clause
\texttt{country\ =\ \textquotesingle{}CH\textquotesingle{}} as a GET
parameter) lists sites in Switzerland. More complex queries can be
executed using nested parameters. For example,
\url{https://xronos.ch/c14s?sample\%5Bmaterial\%5D\%5Bname\%5D=charcoal}
(encoding that the \texttt{c14} table should be joined to the
\texttt{material} table via \texttt{sample}, followed by the WHERE
clause
\texttt{material.name\ =\ \textquotesingle{}charcoal\textquotesingle{}})
lists radiocarbon dates obtained from charcoal samples. Index actions
can also respond with the result in a tabular data format
(i.e.~\texttt{.csv}).

@@

Once ingested, we apply a number of automated and semi-automated quality
control processes to integrate new data into the existing database.
Controlled vocabularies are used in a number of places in the data model
(Figure~\ref{fig-data-model}), and we use thesauri to automatically
standardise these fields as much as possible. For example, as mentioned
above, the taxonomic description of samples is controlled using GBIF's
backbone taxonomy (GBIF Secretariat, 2023), and we also use a thesaurus
service provided by GBIF to automatically change variant or obsolete
taxonomic names to the canonical version. If the system is not able to
standardise a field using the available thesaurus, it is flagged for
manual correction. A wide variety of other potential data quality issues
(e.g.~missing data on what country a site is in) are also flagged for
human review by this system Table~\ref{tbl-issues}, which can often be
semi-automated (e.g.~suggesting close matches in the thesaurus or the
country indicated by the record's coordinates).

A final critical component of XRONOS' data curation system is duplicate
handling. We import data from many overlapping resources (many of which
incorporate each other either in whole or in part), so duplicate records
are common (as recently discussed by Reiter et al., 2024). The end
result of standardising and correcting a record is also often to create
a duplicate: e.g.~the same sample imported from one source as `oak' but
another as `\emph{Quercus} sp.' will become a duplicate pair as
`\emph{Quercus}', and thus be recognised as a single sample. Such exact
duplicates can be merged automatically, with the oldest record becoming
the authoratitive version, but detecting fuzzier duplicated information
(e.g.~differences in the spelling of site names) has proved a more
difficult problem. As of writing there are therefore still many
duplicate records in XRONOS that need to be manually resolved, but we
hope to automate much more of this work in the future.

@@

In this paper we have outlined the conceptual and technical
infrastructure developed during the initial, SNF-funded phases of
development on XRONOS in 2019 and 2021--2024. These include a
generalised data model for radiocarbon and typological dates, extendible
to other chronometric information, and associated contextual
information; an R- and Ruby-based pipeline for continuous ingestion of
data from a variety of sources; continuous, semi-automated data cleaning
protocols; a Ruby-on-Rails application providing a web-based frontend to
the data and a REST API for programmatic access; and an R package for
interfacing with the API. These systems are in production and publicly
accessible at \url{https://xronos.ch}.

At the current stage of implementation, we argue XRONOS provides a
framework for access to chronometric data that is more open, more
reliable, and more comprehensive than the previously available global
radiocarbon compilations. XRONOS blends aspects of the three existing
approaches (c14bazAAR, IntChron, and p3k14c; see Section
Section~\ref{sec-global-compilations}) to achieve the same aim of
providing access to the global radiocarbon data through a common
interface. Like c14bazAAR and IntChron, it is a `metadatabase' that
draws from existing data resources and maintains an explicit link to
them. But like p3k14c, it integrates these into a single database and
applies data curation processes to harmonise them and improve the
quality of the information. It has a wider scope than c14bazAAR or
IntChron, as it mirrors rather than directly retrieves the source data
(allowing us to use resources that aren't openly published), and does
not rely on the authors of these sources to implement a common
specification. It also goes beyond the functionality of p3k14c by
providing systems for the continuous ingestion and curation of new data
as it is published. With that said, we see the approaches followed by as
complementary rather than competing. A c14bazAAR parser for XRONOS is in
development (\url{https://github.com/ropensci/c14bazAAR/pull/150}), and
we also aim to provide an IntChron interface to XRONOS' data in the near
future. New data and corrections from p3k14c are incorporated into
XRONOS as they are released.

@@

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  number,
  preprint]{elsarticle}

@@

\affiliation[1]{organization={University of Bern, Institute of
Archaeological Sciences},,postcodesep={}}
\affiliation[2]{organization={Max Planck Institute for Evolutionary
Anthropology},,postcodesep={}}

@@

In the remainder of this paper, we describe the concept and
implementation of XRONOS in relation to the state of the art in open
chronometric data in archaeology, and evaluate our progress in achieving
these goals as of writing. Since we envisage both XRONOS as a dataset
and XRONOS as software to be continually developing resources, the
description here should be read as a `snapshot' of the project as of
version 1.2.0.

@@

The number of available compilations has increased signifcantly since
around 1995 (Figure~\ref{fig-c14-datasets-time}). The first generation
came around the turn of the century and consists mostly of online
databases with a web frontend. These include some databases operated by
radiocarbon labs, for example the Oxford Radiocarbon Lab (ORAU) and the
Belgian Royal Institute for Cultural Heritage (KIK-IRPA), and
essentially represent a continuation of their date lists in a digital
format. The majority, however, were compiled from the literature by
individual researchers interested in a particular region and/or period.
Notable early examples include ANDES 14C in 1994 (Central Andes,
Michczyński et al., 1995), CARD (Canada, Gajewski et al., 2011) and
RADON (Europe, Raetzel-Fabian, 1999) in 1999, and CANEW in 2001 (Near
East, Reingruber and Thissen, 2005). From 2010, coinciding with broader
shifts in scientific publishing (Tenopir et al., 2011), it became more
common to publish standalone `open data' products in the form of journal
supplements, archives in repositories and/or data papers; the
\emph{\href{https://openarchaeologydata.metajnl.com}{Journal of Open
Archaeology Data}}, launched in 2012, has been a prominent venue for
this latter category. Most recently there has been a trend towards
providing version-controlled plain text data via platforms such as
\href{https://github.com}{GitHub}, reflecting the broader adoption of
these tools amongst computational archaeologists over the last decade
(Batist and Roe, 2024). The shift from online databases towards more
static but more preservable open data products is welcome, given how
many databases from the first generation have subsequently ceased to be
accessible. Version-controlled repositories are particular well-suited
to data compilation projects because they allow for continued updates
whilst still providing snapshot `releases' that are citeable and can be
archived in long-term repositories.

@@

\caption{\label{fig-c14-datasets-map}Geographic coverage of published
regional radiocarbon compilations according to our survey (binned by UN
M49 macroregion).}

@@

\caption{\label{fig-c14-global}Geographic (kernel density) and temporal
(sum calibration) of georeferenced dates in XRONOS and other global
radiocarbon compilations}

@@

Radiocarbon has been by far the most active area of open data
compilation, but archaeological chronology incorporates a much more
diverse range of sources of information (Harding, 1999). In periods
beyond the practical limit of radiocarbon dating (c.~55,000 BP), other
types of radiometric (K--Ar, U--Pb, etc.), chemical or luminescence
dating offer an alternative (Aitken, 1999). Conversely, in historic
periods, or on calibration `plateaus' (Guilderson et al., 2005),
radiocarbon is often less important than conventional typological dating
(based on artefact characteristics), which in these periods can offer
comparable or better temporal resolution, or direct dating based on
epigraphy (Heřmánková et al., 2021), numismatics (Kemmers and Myrberg,
2011) or historical sources. In places where it is widely available,
dendrochronology (Baillie, 2014) also produces significantly better
resolved chronologies and therefore tends to be the main source of
chronometric data. Other more application-specific chronological methods
include shoreline dating (Brøgger, 1905; Roalkvam, 2023), lichenometry
(Benedict, 2009) and rock weathering dating (Bednarik, 2020; Whitley,
2012).

@@

DendroDB, hosted by the Swiss Federal Institute for Forest, Snow and
Landscape Research (WSL), emphasizes ecological and climate studies over
archaeological wood material. The CFS-TRenD database, managed by the
Canadian Forest Service, compiles over 4,600 datasets from Canadian
forests, primarily focusing on boreal ecosystems. Despite its extensive
coverage for North America, its geographical specificity and lack of
open access restrict its utility for European prehistoric contexts.
Similarly, the QUB Dendrochronology Database, managed by Queen's
University Belfast, offers valuable datasets for Ireland and the UK but
lacks significant representation of prehistoric material, limiting its
application in broader archaeological investigations. The Building
Archaeology Research Database (BARD) contains over 24,000 records,
including dendrochronological data from more than 2,700 buildings.
However, its focus on medieval and post-medieval timber-frame
construction further narrows its utility for studies involving
prehistoric wood samples.

@@

Typological dates---i.e.~relative, expertise- or seriation-based dating
based on artefact characteristics---are ubiquitous in archaeological
studies but rarely treated as a form of chronometric data in their own
right. For example, the majority of the radiocarbon datasets we reviewed
(Section~\ref{sec-c14-compilation}) included some form of typological
chronological information in the form of a `period' or `culture' column.
This is also typically present in many other forms of systematic
compilation work in archaeology, for example site gazetteers. Aggregated
typological information from such sources are often used in aoristic
analysis and related methods (Crema, 2024; Mischka, 2004). What is
lacking in this presentation of typological dating is metadata on how
the determination was made and how exactly it is to be understood. Like
any archaeological date, a typological date is derived from a physical
sample -- the object or set of object from which a chronological
estimate was derived. Typological dates on one class of object may well
clash with other classes of object, or for that matter with scientific
dates -- does one trust the date on pottery, the date on architecture,
or the radiocarbon date? Without additional metadata on e.g.~who made
the typological determination or what the radiocarbon date was obtained
on, such inconsistencies are difficult to resolve. Similarly the
absolute date range corresponding to a typological determination
(e.g.~``Late Neolithic'') can be interpreted in multiple ways depending
on the region and intentions of the expert making the determination.
PeriodO (Rabinowitz et al., 2016) is a linked open data infrastructure
that includes a shared vocabulary of typological periods and
corresponding calendar age estimates, and an important step towards
addressing the latter problem. However, it remains to be systematically
linked to compilations of typological dates (though there are some
efforts in this direction e.g. Hannah et al., 2022).

@@

XRONOS is our answer to Kintigh's call (Kintigh, 2006) for digital
infrastructures that don't just provide access to chronological data but
enables researchers to ``archive, access, integrate, and mine disparate
data sets''. It complements several similar open data infrastructures
within and outwith archaeology, such as the Global Biodiverisity
Information Facility (GBIF, Canhos et al., 2004), the Strategic
Environmental Archaeology Database (SEAD, Buckland, 2014), IMPACT for
mummified human remains (Nelson and Wade, 2015), Neotoma for
palaeoecological data (Williams et al., 2018), IsoArcH for stable
isotope data (Plomp et al., 2022), Poseidon for archaeogenetic data
(Schmid et al., 2024), the International Soil Radiocarbon Database
(ISRaD, Lawrence et al., 2020), and the `Big Interdisciplinary
Archaeological Database' (BIAD), an ambitious new initiative to combine
many of these individual domains, including chronology (Reiter et al.,
2024). To improve upon existing global syntheses of radiocarbon dates
(see Section~\ref{sec-global-compilations}), we aimed to develop a
living infrastructure that both continually collected data from diverse
sources and presented a seamless single database to the user.

@@

Chrons are conceptually useful because they emphasise that different
types of archaeological `dates', drawn from different sources, have
essentially the same information content: the location of an event in
space and time. We thereby avoid privileging certain sources of
chronological data over others (as might be the case if, for example, we
treated `period' as a fixed attribute of a site) and can accommodate
contradictory (e.g.~differences of opinion on typological
classification). This is important given that XRONOS aspires to be an
authoritative `backbone' with a global scope, so we cannot realistically
impose a single chronological scheme or resolve conflicting information
provided by specialists. Chrons are useful practically because they
expose a common interface for attributes that all types of chronological
information share, such as a \emph{terminus post quem} (TPQ),
\emph{terminus ante quem} (TAQ), and midpoint estimate. This allows
applications to query chronological data from multiple sources, without
necessarily having to be aware of the pecularities of each type of
dating.

@@

The series of relations
\texttt{(chron)\ \textgreater{}\ sample\ \textgreater{}\ context\ \textgreater{}\ site}
links the chronological and contextual sides of the XRONOS data model.
Each step is a many-to-one association, meaning for example that it is
possible to attach multiple chrons to the same sample (e.g.~replicated
radiocarbon dates on the same material), multiple \emph{types} of chrons
to the same sample (e.g.~radiocarbon dates on tree-rings for
wiggle-matching). Since this kind of information is rarely
systematically recorded in our source databases, there are currently few
actual records that make use of this feature of the data model. However,
we hope it will provide a foundation for more nuanced chronological
modelling in the future.

@@

The XRONOS web application exposes two distinct user interfaces: a
graphical user interface accessed through a web browser; and an
application programming interface (API). Both interfaces follow a REST
(Representational State Transfer) pattern (Verborgh et al., 2015), where
each resource (e.g.~a single radiocarbon date, a single bibliographic
reference, or a single user) is statelessly mapped to a single address.
Users can then interact with resources at these addresses using a
preditable and uniform interface based on HTTP verbs. For example, the
radiocarbon date RTD-8904 is represented by the address
\url{https://xronos.ch/c14s/156205}. Users can view information on this
resource by sending a GET request to that address, regardless of which
interface they are using, and authorised users can modify it using POST,
PATCH, or DELETE requests. The bibliographic reference associated with
this date (Richter et al., 2017) is similarly represented at the address
\url{https://xronos.ch/references/17778}, and can be accessed at that
address using the same interface as the radiocarbon date.

This basic REST pattern is augmented by seven `actions' (following the
standard pattern in Rails application) that express different ways of
interacting with a resource: index, show, destroy, new, create, edit,
and update. The `show' action represents interaction with a single
resource, as described above. The `index' action, which lists resources
of a given type (e.g.~\url{https://xronos.ch/c14s} for radiocarbon
dates), is worth special mention because it is through this that the
filtering logic at the core of XRONOS' two interfaces is implemented. By
passing a query as HTTP GET parameters to the index action of a
resource, the list returned to the user is modified to only include
records that match that query. For example,
\url{https://xronos.ch/sites?site\%5Bcountry_code\%5D=CH} (the part of
the URL after the \texttt{?} character encodes the SQL WHERE clause
\texttt{country\ =\ \textquotesingle{}CH\textquotesingle{}} as a GET
parameter) lists sites in Switzerland. More complex queries can be
executed using nested parameters. For example,
\url{https://xronos.ch/c14s?sample\%5Bmaterial\%5D\%5Bname\%5D=charcoal}
(encoding that the \texttt{c14} table should be joined to the
\texttt{material} table via \texttt{sample}, followed by the WHERE
clause
\texttt{material.name\ =\ \textquotesingle{}charcoal\textquotesingle{}})
lists radiocarbon dates obtained from charcoal samples. Index actions
can also respond with the result in a tabular data format
(i.e.~\texttt{.csv}).

@@

Once ingested, we apply a number of automated and semi-automated quality
control processes to integrate new data into the existing database.
Controlled vocabularies are used in a number of places in the data model
(Figure~\ref{fig-data-model}), and we use thesauri to automatically
standardise these fields as much as possible. For example, as mentioned
above, the taxonomic description of samples is controlled using GBIF's
backbone taxonomy (GBIF Secretariat, 2023), and we also use a thesaurus
service provided by GBIF to automatically change variant or obsolete
taxonomic names to the canonical version. If the system is not able to
standardise a field using the available thesaurus, it is flagged for
manual correction. A wide variety of other potential data quality issues
(e.g.~missing data on what country a site is in) are also flagged for
human review by this system (Table~\ref{tbl-issues}), which can often be
semi-automated (e.g.~suggesting close matches in the thesaurus or the
country indicated by the record's coordinates).

A final critical component of XRONOS' data curation system is duplicate
handling. We import data from many overlapping resources (many of which
incorporate each other either in whole or in part), so duplicate records
are common (as recently discussed by Reiter et al., 2024). The end
result of standardising and correcting a record is also often to reveal
a duplicate: e.g.~the same sample imported from one source as `oak' but
another as `\emph{Quercus} sp.' will become a recognisable duplicate
record as `\emph{Quercus}', and thus can be automatically merged to a
single sample. Such exact duplicates can be merged automatically, with
the oldest record becoming the authoratitive version, but detecting
fuzzier duplicated information (e.g.~differences in the spelling of site
names) has proved a more difficult problem. As of writing there are
therefore still many duplicate records in XRONOS that need to be
manually resolved, but we hope to automate much more of this work in the
future.

@@

In this paper we have outlined the conceptual and technical
infrastructure developed during the initial, SNF-funded phases of
development on XRONOS in 2019 and 2021--2024. These include a
generalised data model for radiocarbon and typological dates, extendible
to other chronometric information, and associated contextual
information; an R- and Ruby-based pipeline for continuous ingestion of
data from a variety of sources; continuous, semi-automated data cleaning
protocols; a Ruby-on-Rails application providing a web-based frontend to
the data and a REST API for programmatic access; and an R package for
interfacing with the API. These systems are deployed and publicly
accessible at \url{https://xronos.ch}.

At the current stage of implementation, we argue XRONOS provides a
framework for access to chronometric data that is more open, more
reliable, and more comprehensive than the previously available global
radiocarbon compilations. XRONOS blends aspects of the three existing
approaches (c14bazAAR, IntChron, and p3k14c; see Section
Section~\ref{sec-global-compilations}) to achieve the same aim of
providing access to the global radiocarbon data through a common
interface. Like c14bazAAR and IntChron, it is a `metadatabase' that
draws from existing data resources and maintains an explicit link to
them. But like p3k14c, it integrates these into a single database and
applies data curation processes to harmonise them and improve the
quality of the information. It has a wider scope than c14bazAAR or
IntChron, as it mirrors rather than directly retrieves the source data
(allowing us to use resources that aren't openly published), and does
not rely on the authors of these sources to implement a common
specification. It also goes beyond the functionality of p3k14c by
providing systems for the continuous ingestion and curation of new data
as it is published. With that said, we see the approaches followed by as
complementary rather than competing. A parser for XRONOS has been
contributed back to c14bazAAR and we also aim to provide an IntChron
interface to XRONOS' data in the near future. New data and corrections
from p3k14c are incorporated into XRONOS as they are released.

@@

\bibitem[\citeproctext]{ref-GuldersonEtAl2005}
Guilderson, T.P., Reimer, P.J., Brown, T.A., 2005. The {Boon} and {Bane}
of {Radiocarbon Dating}. Science 307, 362--364.
\url{https://doi.org/10.1126/science.1104164}

@@

\bibitem[\citeproctext]{ref-SchmidEtAl2024}
Schmid, C., Ghalichi, A., Lamnidis, T.C., Mudiyanselage, D.B.A., Haak,
W., Schiffels, S., 2024. Poseidon -- {A} framework for archaeogenetic
human genotype data management. eLife 13.
\url{https://doi.org/10.7554/eLife.98317.1}
